---
title: "Variation in surprisal across seeds and scales"
author: "Pam Rivière and Sean Trott"
date: "March 17, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data


Here, we analyze data looking at the average attention each head gives from each token to the previous token.

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/seed-variability/src/analysis")
directory_path <- "../../data/processed/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) 
nrow(df_pythia_models)

table(df_pythia_models$seed)
table(df_pythia_models$dataset_name)
max(df_pythia_models$step)

### Normalize surprisal by #tokens
df_pythia_models = df_pythia_models %>%
  mutate(surprisal_normed = surprisal / num_tokens)

```


# Surprisal over time by seed

**TODO**:

- Mean/SD surprisal over time, by seed, by architecture. 
- Scaling law in `surprisal ~ #params` over different checkpoints. 
- Do larger models have more or less variance across seeds in surprisal? 
- Are larger models more or less *correlated* at a word-by-word level? 
- Are larger models correlated across seeds more *quickly* throughout pretraining? What about with smaller models, i.e., are larger models already capturing the wisdom of the crowd of smaller models?
(For this, compare the *average* of 14m at a given timepoint with each individual 14m seed, as well as with 70m at that timepoint.)
- How does within-model (across-seed) correlation compare to across-models? When does a model "look like itself"? (MDS, etc.)
- Compare each model to a set of baseline models at each timepoint (e.g., GPT-2), to get a sense for the similarity to another model family.
- How do these trajectories compare to the trajectories of weight changes?
- Also do with Pythia cosine distances for RAW-C?


```{r}
df_surprisal_by_seed = df_pythia_models %>% 
  group_by(n_params, model, seed_name, revision, step, seed) %>%
  summarise(mean_surprisal = mean(surprisal_normed),
            sd_surprisal = sd(surprisal_normed)) %>%
  mutate(step_modded = step + 1)

df_surprisal_avg = df_pythia_models %>% 
  group_by(n_params,model, revision, step) %>%
  summarise(mean_surprisal = mean(surprisal_normed),
            sd_surprisal = sd(surprisal_normed))  %>%
  mutate(step_modded = step + 1)


df_surprisal_by_seed %>%
  ggplot(aes(x = step_modded,
             y = sd_surprisal,
             color = seed_name)) +
  geom_line(size = .7, alpha = .7, linetype = "dotted") +
  geom_line(data = df_surprisal_avg, aes(x = step_modded, y = sd_surprisal), 
            color = "black", size = 1.3) + # Smoothed average
  geom_point(data = df_surprisal_avg, aes(x = step_modded, y = sd_surprisal), 
             color = "black", size = 2) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "SD Surprisal",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~reorder(model, n_params))



### Seed-wise averages
df_surprisal_avg = df_surprisal_by_seed %>% 
  group_by(n_params,model, revision, step) %>%
  summarise(mean_surprisal_across_seeds = mean(mean_surprisal),
            surprisal_var_across_seeds = sd(mean_surprisal),
            surprisal_mean_range = max(mean_surprisal) - min(mean_surprisal),
            se_surprisal = sd(mean_surprisal) /sqrt(n()))  %>%
  mutate(step_modded = step + 1)

ggplot() +
  # Raw points from individual seeds
  geom_point(data = df_surprisal_by_seed,
             aes(x = step_modded,
                 y = mean_surprisal,
                 color = reorder(model, n_params)),
             size = 3, alpha = 0.3) +

  # Summary lines and points
  geom_line(data = df_surprisal_avg,
            aes(x = step_modded,
                y = mean_surprisal_across_seeds,
                color = model),
            size = .7,
            linetype = "dashed") +
  # Formatting
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Mean Surprisal",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                   begin = 0.8, end = 0.15))



df_surprisal_avg %>%
  ggplot(aes(x = step_modded,
             y = surprisal_var_across_seeds,
             color = model)) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "SD Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                begin = 0.8, end = 0.15))



df_surprisal_avg %>%
  ggplot(aes(x = step_modded,
             y = surprisal_mean_range,
             color = model)) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Range in mean surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                begin = 0.8, end = 0.15))


```


# Across-seed correlations

**TODO**: Figure this out...right now code is not dealing well with multiple words per sentence (e.g., "the") when pivoting wider. (Solution: just avg across those instances)


**TODO**: Weird effect where correlation is higher early on...not sure what this means re: initialization? (Slution: it was about multi-token words)


```{r}

#### By model/seed
df_wide <- df_pythia_models %>%
  filter(seed < 9) %>% ### REMOVE
  mutate(model_id = paste(model, "-", seed_name)) %>%
  select(step, model_id, surprisal_normed, sentence, word) %>%
  group_by(model_id, step, sentence) %>%
  summarise(surprisal = mean(surprisal_normed, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = model_id,
    values_from = surprisal
  )


cor_matrix <- df_wide %>%
  filter(step == 16) %>%
  mutate(`pythia-14m - mean` = rowMeans(select(., starts_with("pythia-14m")), na.rm = TRUE),
         `pythia-70m - mean` = rowMeans(select(., starts_with("pythia-70m")), na.rm = TRUE),
         `pythia-160m - mean` = rowMeans(select(., starts_with("pythia-160m")), na.rm = TRUE)) %>%
  select(-step, -sentence) %>%
  cor(use = "pairwise.complete.obs")



# Plot the correlation matrix
ggcorrplot(cor_matrix, 
           hc.order = FALSE,
           method = "square" 
          )



cor_by_group <- df_wide %>%
  group_by(step) %>%
  summarise(
    cor_matrix = list(
      cor(select(cur_data(), starts_with("pythia-")), use = "pairwise.complete.obs")
    ),
    .groups = "drop"
  )


cor_long <- cor_by_group %>%
  mutate(cor_df = map(cor_matrix, ~ as.data.frame(as.table(.)))) %>%
  unnest(cor_df) %>%
  rename(seed1 = Var1, seed2 = Var2, correlation = Freq) %>%
  filter(seed1 != seed2) %>%
  separate(seed1, into = c("model1", "seed_name1"), sep = " - ") %>%
  separate(seed2, into = c("model2", "seed_name2"), sep = " - ") %>%
  mutate(same_model = model1 == model2)

# Step 2: Compute mean correlation per step (and optionally model)
mean_cor_by_step <- cor_long %>%
  group_by(model1, model2, same_model, step) %>%
  summarise(
    mean_corr = mean(correlation, na.rm = TRUE),
    se_corr = sd(correlation, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(mean_cor_by_step, aes(x = step + 1, y = mean_corr, color = model2)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_corr - 2 * se_corr,
                  ymax = mean_corr + 2 * se_corr,
                  fill = model2),
              alpha = 0.5, color = NA) +
  theme_minimal() +
  labs(
    title = "Inter-seed Correlation",
    x = "Training Step (Log10)",
    y = "Mean Correlation",
    color = "",
    fill = ""
  ) +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                   begin = 0.8, end = 0.15)) +
  scale_fill_manual(values = viridisLite::viridis(3, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  facet_wrap(~model1)


summary(lmer(correlation ~ step * same_model + 
               (1 | seed_name1) + (1 | seed_name2), data = cor_long))




##### Multi-dimensional scaling
mds_by_step <- df_wide %>%
  filter(step <= 128) %>%
  drop_na() %>%
  group_by(step) %>%
  summarise(
    mds_df = list({
      cor_mat <- cor(select(cur_data(), starts_with("pythia-")), use = "pairwise.complete.obs")
      dist_mat <- as.dist(1 - cor_mat)
      mds <- cmdscale(dist_mat, k = 2)
      as_tibble(mds, .name_repair = "unique") %>%
        mutate(model_id = colnames(cor_mat)) 
    }),
    .groups = "drop"
  ) %>%
  unnest(mds_df) %>%
  separate(model_id, into = c("model", "seed"), sep = " - ") %>%
  rename(x = `...1`, y = `...2`)

ggplot(mds_by_step, aes(x, y, color = model)) +
  geom_point() +
  facet_wrap(~ step) +
  theme_bw() +
  labs(x = "MDS 1",
       y = "MDS 2",
       color = "") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                   begin = 0.8, end = 0.15)) 


# Create animated plot
p <- ggplot(mds_by_step, aes(x = x, y = y, color = model)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 14) +
  labs(title = "MDS of Model Similarities (Step: {closest_state})",
       x = "MDS Dimension 1",
       y = "MDS Dimension 2",
       color = "Model") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                   begin = 0.8, end = 0.15)) +
  transition_states(step, transition_length = 2, state_length = 1) +
  ease_aes('linear')

# Render
anim <- animate(p, renderer = gifski_renderer(), width = 800, height = 600, 
                fps = 10, duration = 10)
anim


```

